A Bayesian Framework for Learning Regular Expressions

### 2012-12-08 paper/paper.tex
Gneral corrections to the paper.

\textbf{Keywords:} 
Model inference, regular expression.
\end{abstract}

@@

%\subsection{Balancing Generality with Specificity}
Our model has two parts: We generate a set of regular expressions which describe a set of input stings $S$, and we find the regular expression in that set with the highest posterior probability. The posterior probability is proportional to the product of the hypothesis' prior probability and its likelihood.
\begin{align*}
	p(h|S) \propto p(h) p(S|h)
\end{align*}
Below, we define the prior probability such that our model biases smaller hypotheses, and the likelihood of generating a given hypothesis is inversely proportional to its cost. Our model aims to balance generality with specificity. For example, given "ABC, ABD", we wish to favor the regex $AB(C|D)$ over "?*", which is less likely given the size principle. At the same time, given "ABA", "ABC", "ABD", "ABE", "ABF", we may wish to favor 'AB?' over  '$AB(A|C|D|E|F)$'

\subsection{Representation of Hypotheses}
Regular expressions are equivalent to deterministic finite automata (DFA).\footnote{See the Appendix for a brief summary of our regex notation.} Because DFA are often easier to reason with, we use them to represent regular expressions in our model (See Fig 1 for an example DFA.)

@@

In this sense, our DFA is similar to a probabilistic automaton, although the acceptance criteria is $\tau=0$, so the languages represented by our DFA are still regular. [Citation]. A second difference is that there is no concept of stopping string generation in probabilistic automaton.

@@

\subsection{Generating Hypotheses}
The hypothesis space is initiated with the base hypothesis, the term applied to a DFA that accepts strictly the input strings. Next, hypotheses are added to the hypothesis space through two processes of generalization: merging and applying wildcards.

@@

We resolve this issue by performing beam search. In beam first, we perform breadth first search, but for any given depth of the search tree, we only consider children of the $m$ most promising hypotheses. In our case, the most promising hypotheses were defined as those with the highest posterior probability.

@@

\subsection{Prior Probability}
We assigned the prior probability to the hypotheses (i.e. regexes) based on the size of the DFA and the number of transitions. In the following equation, $|S|$ is the number of states in the DFA and $|T|$ is the number of tranisitons in the DFA, where a wildcard counts as only one transition. 
\begin{align*}
	P(h) \propto e^{-\alpha|S| - \beta|T|}
\end{align*}
Thus, we have biased smaller DFA that use wildcards instead of literal characters. [cite Program Merging model]

\subsection{Likelihood}
To define the likelihood of producing certain strings, we turn a DFA into a generative automaton that has a probability of generating a given string. Letting the input strings be independent, the probability of generating strings $S$ is equal to the product of the probabilities of generating each string.
\begin{align*}
	p(S) = \prod_{s \in S} p(s)
\end{align*}

Let the $s$ be the concatenation of characters $a_1a_2...a_n$. We define the probability of generating a string as the product of the probability of generating the letters in sequence $p(a_1a_2...a_n)$ and the product of stopping the string generation on an accept state $p(\text{stop})$.
\begin{align*}
	p(s) = p(a_1a_2...a_n)p(\text{stop})
\end{align*}

We define the probability of generating the characters as the product of the probabilities of making each state transition over each state.
\begin{align*}
	p(a_1a_2 ... a_n) = \prod_{i=1}^{n}p(\delta(r_i, a_{i+1}))
\end{align*}
where $r_i$ is the state reached after the first $i-1$ characters, and $r_1$ is the start state.

To simplify inference, we assume uniform probability over all possible state transitions. Formally, the probability of making a state transition is given by:
\begin{align*}
	p(\delta(r_i, a_{i+1})) &= \frac{1}{\sum_{a\in A} 1\{\delta(r_i, a)\}}
\end{align*}
Notice that the denominator counts the number of characters for which a transition is defined. In the case of a wildcard, the transition is defined for all 26 letters.

@@

The probability of stopping at an accept state is given by the equation below. The condition $\exists a : \delta(r_n, a)$ indicates that there is an outbound transition for the accept state of the string.
\begin{align*}
	p(\text{stop}) &= \left\{
		\begin{array}{rl}
		\tau (1-\tau)^b & r_n \in F \wedge \exists a : \delta(r_n, a)
		\\ (1-\tau)^b & r_n \in F \wedge \not \exists a : \delta(r_n, a)
		\\ 0 & \text{otherwise}
		\end{array}
		\right.
\end{align*}

@@

Our equation for $p(\text{stop})$ is thus:
\begin{align*}
	p(\text{stop}) &= \left\{
		\begin{array}{rl}
		\frac{1}{b^2+3b+1} & \in F \wedge \exists a : \delta(r_n, a)
		\\ \frac{1}{b+1} & \in F \wedge \not \exists a : \delta(r_n, a)
		\\ 0 & \text{otherwise}
		\end{array}
		\right.
\end{align*}

@@

\begin{align*}
	p(s \in \mathcal{L} | S \subseteq {L}) = \sum_{h \in \mathcal{H}} 1 (s \in h) p(h|S)
\end{align*}

For human participants, because the hypothesis is unobserved, we take $p(s \in \mathcal{L} | S \subseteq {L}) $ to be the average acceptance of a test string given a set $S$.

In our model, we consider $\mathcal{H}$ to be the set of all hypotheses generated in our model.

\section{Results and Discussion}
Figure 3 shows the comparison between our model and the human subject data.

\begin{figure}[ht]
\begin{center}
\begin{align*}
\includegraphics[scale=0.6]{results.png}
\end{align*}
\end{center}
\caption{Chart showing  $p(s \in \mathcal{L} | S \subseteq {L}) $. Human subject data is in red, model data is in blue. $R$ = .75} 
\label{merge-figure}
\end{figure}

@@

We note that in cases 1 (S = \{'yyyt', 'yt', 'yyyt', 'yyyt', 'yt'\}, s= 'yyyyyt' ), 4( S = \{jojojojo', 'jojojojo', 'jojojojo'\}, s = jojo), and 6 (s =  tw,  S = \{'twtwtw', 'twtwtw', 'twtwtw'\}), it is clear that the model weighs a smaller number of transitions greater than human subjects do. However, because this same parameter in part controls for wildcards, lowering it reduces performance on many other cases. This suggests that our model's parameter $\beta$ may in fact be better represented by more than one variable in human decision making.

We also observe that in such cases, there was a relatively small set of probable hypotheses, and that any favoring of smaller hypotheses was magnified by this fact. 
\section{Conclusion and Future Work}

Despite the number of ways in which our model might be improved mentioned above,  it remains highly implausible that the brain generates its set of candidate hypotheses by ever considering even $2^n$ hypotheses in the first level of beam search we perform, much less far more than that. Doing so would be highly inefficient: the number of implausible hypotheses generated would in most cases be far greater than the number of `good' hypotheses. Far more plausible a priori is the hypothesis that the brain uses of number of heuristics to generate a certain set of candidate hypothesis, and then uses some process analogous to Bayesian inference to select among those hypotheses. To shed light on how this hypothesis selection process may occur, we designed an experiment in which candidates chose among a number of possible rules to explain a set of strings S, but have yet to gather enough data to reach any significant conclusions.

Further experiments could be conducted to determine what heuristics may be at work in the hypothesis generation process. For any set of candidate heuristics, it will remain to be determined whether or not such heuristics are at some point learned. Clearly some are when semantic considerations are at work: cat and mat but not qat are meaningful english words; 'ao' and 'eo' and 'io' begin with vowels unlike 'xo'. This experiment was designed to avoid such candidate hypotheses.

@@

\section{Appendix: Technical Notes}
\subsection{Regex and DFA Notation}
We used a limited alphabet wherein normal characters are all lower case. We also used the following rules to define regular expressions.
\begin{enumerate}
	\item Acceptance of specific strings is represented as concatenated characters. E.g. \verb!abc! accepts only \verb!abc!.
	\item Acceptance of disjunctive strings is represnted by \verb![a|b]!. E.g. \verb![abc|c]!accepts only \verb!abc! and \verb!c!.
	\item Acceptance of any character is represented by the wildcard \verb!?!. E.g. \verb!a?! accepts \verb!aa!, \verb!ab!, \verb!ac!, etc.
	\item Acceptance of the empty string is represented by \verb!E!. \verb!Ea! is the same as \verb!a!.
	\item Acceptance of any sequence repeated zero or more times is represented by the Kleene star, \verb!a*!. E.g. \verb!a*! accepts \verb!E!, \verb!a!, \verb!aa!, etc, 
\end{enumerate}

% \subsection{Sizing the Hypothesis Space}
% Ignoring wildcards and assuming that no chain-reaction merges occur, each hypothesis can be defined as a set of mutually exclusive subsets of the states contained in the base hypothesis. For instance, given states $s_1, s_2, s_3$, the possible hypotheses are:
% 	\begin{align*}
% 		&s_1, s_2, s_3
% 		\\&\text{merge}(s_1, s_2), s_3
% 		\\&\text{merge}(s_1, s_3), s_2
% 		\\&\text{merge}(s_2, s_3), s_1
% 		\\&\text{merge}(s_1, s_2, s_3)
% 	\end{align*}
% To generalize this to a case of $n_b$ states in the base hypothesis, one can consider individually the number of regexes with $n_f$ states in the finished regex. For the cases where $n_b=n_f$ or $n_f=1$, it is easy to verify that there exist only one regex. When creating hypotheses where $1 < n_f < n_b$, however, one can simply look to the case of $n_b - 1$. In particular, for every $n_f$, one can simply duplicate all of the hypotheses from the $n_b - 1$ case with an additional unmerged state $s_{n_b}$. Further, one can add a new hypothesis for each subset of the hypotheses in the $n_s-1, n_f$ case. So, the expression for the total number of regexes given the number of states in the base hypothesis is given by $g(n_b)$.
% 	\begin{align*}
% 	f(a, b) &= \left\{
% 			\begin{array}{cl}
%          		1 & a=b \vee b=1 \\
%          		f(a-1, b-1) + b f(a-1, b) & \mbox{otherwise}
%             \end{array}\right.
% 		\\ g(n_b) &=\sum_{n_f=1}^{n_b} f(n_b, n_f)
% 	\end{align*}
% One of our experiment examples contained 25 nodes in the base hypothesis, which means that the hypothesis space is $g(25)\approx 5.0 \times 10^{17}$ regular expressions. Needless to say, it is intractable to exhaustively search the hypothesis space.

>>>>

\textbf{Keywords:} 
Model inference, regular language, regular expression, deterministic finite automaton.
\end{abstract}

@@

%\subsection{Balancing Generality with Specificity}
Our model has two parts: We generate a set of regular expressions which describe a set of input stings $S$, and we find the regular expression in that set with the highest posterior probability. The posterior probability is proportional to the product of the hypothesis' prior probability and its likelihood.

Below, we define the prior probability such that our model biases smaller hypotheses, and the likelihood of generating a given hypothesis is inversely proportional to its cost. Our model aims to balance generality with specificity. For example, given \verb!abc, abd!, we wish to favor the regex \verb!ab(c|d)! over \verb!A*! where \verb!A! represents any letter,\footnote{See the Appendix for a brief summary of our regex notation.} which is less likely given the size principle. At the same time, given \verb!aba!, \verb!abc!, \verb!abd!, \verb!abe!, \verb!abf!, we may wish to favor \verb!abA! over \verb!ab(a|c|d|e|f)!.

\subsection{Representation of Hypotheses}
Regular expressions are equivalent to deterministic finite automata (DFA). Because DFA are often easier to reason with, we use them to represent regular expressions in our model (See Figure~\ref{merge-figure} for an example DFA.)

@@

Our generative version of the DFA is not to be confused with a probabilistic atumaton, which has transition probabilities much as described above and a final acceptance threshold. Our generative automata does not have this threshold, and instead has an added parameter which controls the likelihood of stopping to generate a string.

As our alphabet, we use the set of all lowercase letters. This set, we feel, has an intuitive boundary, as subjects seeing strings of all lowercase letters are unlikely to suppose an arbitrary uppercase letter or special character, and thus we can assume fixed the expressiveness of wildcards.

@@

\subsection{Generating Hypotheses}
The hypothesis space is initiated with a DFA that accepts strictly the input strings, which we call the \emph{base hypothesis}. Next, hypotheses are added to the hypothesis space through two processes of generalization: merging and applying wildcards.

@@

We resolve this issue by performing beam search. In beam search, we perform breadth first search, but for any given depth of the search tree, we only consider children of the $m$ most promising hypotheses. In our case, the most promising hypotheses were defined as those with the highest posterior probability.

@@

\subsection{Prior Probability}
We assigned the prior probability to the hypotheses (i.e. regexes) based on the size of the DFA and the number of transitions. In the following equation, $|S|$ is the number of states in the DFA and $|\delta|$ is the number of tranisitons in the DFA, where a wildcard counts as only one transition. 
\begin{align*}
	P(h) \propto e^{-\alpha|S| - \beta|\delta|}
\end{align*}
Thus, we have biased smaller DFA that use wildcards instead of literal characters. [cite Program Merging model]

\subsection{Likelihood}
To define the likelihood of producing certain strings, we turn a DFA into a generative automaton with a probability distribution over generated strings. Letting the input strings be independent, the probability of generating the set of strings $S$ is equal to the product of the probabilities of generating each string.
\begin{align*}
	p(S) = \prod_{s \in S} p(s)
\end{align*}

Let the string $s$ be the concatenation of characters $\{a_0,a_1,...,a_{n-1}\}$. We define the probability of generating a string as the product of the probability of generating the letters in sequence $p(a_0a_1...a_{n-1})$ and the product of stopping the string generation on an accept state $p(\text{stop})$.
\begin{align*}
	p(s) = p(a_1a_2...a_n)p(\text{stop})
\end{align*}

We define the probability of generating the characters as the product of the probabilities of making each state transition.
\begin{align*}
	p(a_1a_2 ... a_n) = \prod_{i=0}^{n-1}p(\delta(r_i, a_i))
\end{align*}
where $r_i$ is the state reached after the first $i$ characters, and $r_0$ is the start state.

To simplify inference, we assume uniform probability over all possible outbound transitions for a given state. Formally, the probability of making a state transition is given by:
\begin{align*}
	p(\delta(r_i, a_{i+1})) &= \frac{1}{\sum_{a\in A} 1\{\delta(r_i, a)\in Q\}}
\end{align*}
where $1\{\cdot\}$ is the indicator function, which evaluates to $1$ if the enclosed proposition is true and $0$ otherwise. $\delta(r_i, a)\in Q$ indicates that the state transition leads to another state of the DFA, so the denominator counts the number of characters for which a transition is defined. In the case of the wildcard \verb!A!, the transition is defined for all 26 letters, so the denominator is 26.

@@

The probability of stopping at an accept state is given by the equation below. The condition $r_n \in F \wedge \exists a : \delta(r_n, a)\in Q$ states that the last state reached by the string is an accept state, and that that state has at least one outbound transition.
\begin{align*}
	p(\text{stop}) &= \left\{
		\begin{array}{rl}
		\tau (1-\tau)^b & r_n \in F \wedge \exists a : \delta(r_n, a)\in Q
		\\ (1-\tau)^b & r_n \in F \wedge \not \exists a : \delta(r_n, a)\in Q
		\\ 0 & \text{otherwise}
		\end{array}
		\right.
\end{align*}

@@

Our equation for $p(\text{stop})$ is thus:
\begin{align*}
	p(\text{stop}) &= \left\{
		\begin{array}{rl}
		\frac{1}{b^2+3b+1} & r_n \in F \wedge \exists a : \delta(r_n, a)
		\\ \frac{1}{b+1} & r_n \in F \wedge \not \exists a : \delta(r_n, a)
		\\ 0 & \text{otherwise}
		\end{array}
		\right.
\end{align*}

@@

\begin{align*}
	p(s \in \mathcal{L} | S \subseteq {\mathcal{L}}) = \sum_{h \in \mathcal{H}} 1 \{s \in h\} p(h|S)
\end{align*}

For human participants, because the hypothesis is unobserved, we take $p(s \in \mathcal{L} | S \subseteq {\mathcal{L}}) $ to be the average acceptance of a test string given a set $S$.

We define $\mathcal{H}$ to be the set of all hypotheses generated in our model. Note, however, that this is the set of regular languages which were identified in beam search, so it does not represent the set of all possible hypotheses.

\section{Results and Discussion}
Figure~\ref{results-chart} shows the comparison between our model and the human subject data.

\begin{figure}[ht]
\begin{center}
\begin{align*}
\includegraphics[scale=0.6]{results-1.png}
\end{align*}
\end{center}
\caption{Chart showing  $p(s \in \mathcal{L} | S \subseteq {L}) $. Human subject data is in red, model data is in blue. $R$ = .75} 
\label{results-chart}
\end{figure}

@@

We note that in cases 1 (S = \{\verb!yyyt!, \verb!yt!, \verb!yyyt!, \verb!yyyt!, \verb!yt!\}, s= \verb!yyyyyt!), 4 (S = \{\verb!jojojojo!, \verb!jojojojo!, \verb!jojojojo!\}, s = \verb!jojo!), and 6 (S = \{\verb!twtwtw!, \verb!twtwtw!, \verb!twtwtw!\}, s =  \verb!tw!), it is clear that the model weighs a smaller number of transitions greater than human subjects do. However, because this same parameter in part controls for wildcards, lowering it reduces performance on many other cases. This suggests that our model's parameter $\beta$ may in fact be better represented by more than one variable in human decision making.

We also observe that in such cases, there was a relatively small set of probable hypotheses, and that any favoring of smaller hypotheses was magnified by this fact. 
\section{Conclusion and Future Work}
future work idea:

-different wildcards (numbers, etc). this was implemented, but we didn't implement any experiments

@@

\section{Appendix: Technical Notes}
\subsection{Regex and DFA Notation}
We used a limited alphabet wherein normal characters are all lower case. We also used the following rules to define regular expressions.
\begin{enumerate}
	\item Acceptance of specific strings is represented as concatenated characters. E.g. \verb!abc! accepts only \verb!abc!.
	\item Acceptance of disjunctive strings is represnted by \verb![a|b]!. E.g. \verb![abc|c]!accepts only \verb!abc! and \verb!c!.
	\item Acceptance of any character is represented by the wildcard \verb!A!. E.g. \verb!aA! accepts \verb!aa!, \verb!ab!, \verb!ac!, etc.
	\item Acceptance of the empty string is represented by \verb!E!. \verb!Ea! is the same as \verb!a!.
	\item Acceptance of any sequence repeated zero or more times is represented by the Kleene star, \verb!a*!. E.g. \verb!a*! accepts \verb!E!, \verb!a!, \verb!aa!, etc, 
\end{enumerate}

\subsection{Sizing the Hypothesis Space}
Ignoring wildcards and ignoring the mergings that require subset construction, each hypothesis can be defined as a set of mutually exclusive subsets of the states contained in the base hypothesis. For instance, given states $s_1, s_2, s_3$, the possible hypotheses are:
	\begin{align*}
		&s_1, s_2, s_3
		\\&\text{merge}(s_1, s_2), s_3
		\\&\text{merge}(s_1, s_3), s_2
		\\&\text{merge}(s_2, s_3), s_1
		\\&\text{merge}(s_1, s_2, s_3)
	\end{align*}
To generalize this to a case of $n_b$ states in the base hypothesis, one can consider individually the number of regexes with $n_f$ states in the finished regex. For the cases where $n_b=n_f$ or $n_f=1$, it is easy to verify that there exist only one regex. When creating hypotheses where $1 < n_f < n_b$, however, one can simply look to the case of $n_b - 1$. In particular, for every $n_f$, one can duplicate all of the hypotheses from the $n_b - 1$ case with an additional unmerged state $s_{n_b}$. Further, one can add a new hypothesis for each subset of the hypotheses in the $n_s-1, n_f$ case. So, the expression for the total number of regexes given the number of states in the base hypothesis is given by $g(n_b)$.
	\begin{align*}
	f(a, b) &= \left\{
			\begin{array}{cl}
         		1 & a=b \vee b=1 \\
         		f(a-1, b-1) + b f(a-1, b) & \mbox{otherwise}
            \end{array}\right.
		\\ g(n_b) &=\sum_{n_f=1}^{n_b} f(n_b, n_f)
	\end{align*}
One of our experiment examples contained 25 nodes in the base hypothesis, which means that the hypothesis space is $g(25)\approx 5.0 \times 10^{17}$ regular expressions. Needless to say, it is intractable to exhaustively search the hypothesis space.
